---
title: "Tutorial 03: Simular Datos"
author: "Grace Smith-Vidaurre"
date: "2023-12-27"
output: 
  html_document:
    css: "styles.css"
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

```

<h2>Información sobre esta traducción</h2>

Este tutorial fue traducido al español por Grace Smith-Vidaurre, siguiendo las [convenciones de traducción de los Carpentries](https://github.com/Carpentries-ES/board/blob/master/Convenciones_Traduccion.md), incluyendo usar el género femenino por defecto. Si encuentras errores de ortografía que impiden tu habilidad de completar los tutoriales, por favor reporta los errores de ortografía a GitHub usando los pasos arriba para reportar un "Issue".

<h2>Resumen del tutorial y objetivos de aprendizaje</h2>

En este tercer tutorial, vamos a crear datos simulados de moviemientos de animals que fueron detectados por diferentes sensores. El proceso de simular estos datos reemplaza el proceso de coleccion de datos que provee ABISSMAL para grabar datos de animales en vivo. Generar estos datos simulados te va a proveer mas oportunidades para practicar habilidades basicas de escribir codigo y tener control sobre la creacion de estos datos te ayudara entender los pasos diferentes de analisis de datos que siguen. Si quieres ver datos recolectados de pajaros con el software de ABISSMAL, y el codigo que fue usado para analizar esos datos, puedes revisar el preprint del [manuscrito de metodos](https://ecoevorxiv.org/repository/view/6268/) que tiene enlaces a los datos y el codigo que son publicamente acesibles.

A traves del proceso de simular datos en este tutorial, vas a continuar usando habilidades de programacion que aprendiste en el segundo tutorial, y vas a aprender habilidades adicionales que incluyen:

1. Crear objetos como vectores y `dataframes`
2. Tipos de datos en R
3. Indexar y manipular objetos
4. Usar frases condicionales
5. Expresiones de `pipe` en el `tidyverse`

<h2>Cargar paquetes</h2>

En el tutorial anterior instalaste el `tidyverse`, una coleccion de paquetes para la ciencia de datos. Tambien aprendiste sobre directorios de trabajo y creaste un directorio nuevo en tu computadora para guardar archivos de datos o imagenes que vas a generar en los siguientes tutoriales.

Cada vez que inicies un archivo nuevo de RMarkdown o R, es importante configurar tu espacio virtual antes de empezar a analizar datos. En el trozo abajo, vas a limpiar tu ambiente global y cargar el `tidyverse` usando codigo que viste en el segundo tutorial, pero ahora todo el codigo esta combinado en un solo trozo.
```{r message = FALSE, warning = FALSE}

rm(list = ls()) # Limpia tu ambiente global

library(tidyverse) # Carga la coleccion de paquetes en el tidyverse

```

<h2>Crear un objeto de `path`</h2>

El siguiente paso sera especificar tu directorio de trabajo. En el segundo tutorial, usaste un `string`, o una secuencia de caracteres entrecomillas para indicar texto adentro de codigo de R, para especificar el `path` de tu directorio de trabajo mientras usabas funciones diferentes. En vez de copiar y pegar la misma secuencia de caracteres cada vez que quieres usar este `path`, es mas eficiente guardar esta secuencia de caracteres (que especifica el `path` de tu directorio de trabajo) adentro de un objeto nuevo, y luego usar el nombre del objeto cuando necesitas especificar el `path`.

Para crear un objeto de tu `path`, puedes escribir el nombre del objeto que quieres crear al lado izquierdo del trozo (sin comillas), luego los simbolos para crear un objeto en R (`<-`), y luego la informacion que quieres asignar a este objeto. En este case, la informacion que quieres guardar adentro de este objeto es tu directorio de trabajo en el formato de un `string`, y esta informacion necesita estar entrecomillas.
```{r eval = TRUE}

path <- "/home/gsvidaurre/Desktop/ABISSMAL_tutoriales"

```

En el codigo arriba, creaste un objeto que se llama `path`. Puedes ver la informacion que contiene este objeto con escribir el nombre del objeto y ejecutar ese codigo en la consola:
```{r eval = TRUE}

path

```

Tambien puedes ver el contenido de `path` con hacer clic en la pestana de "Environment" y revisar la columna al lado derecho del nombre del objeto. Ahora puedes confirmar que `path` es un objeto nuevo que contiene informacion sobre tu directorio de trabajo y esta disponible en tu ambiente global para mas operaciones.

Puedes practicar eliminar solamente el objeto `path` de tu ambiente global con escribir y ejecutar el codigo `rm(list = "path")`. Despues de ejecutar este codigo, deberias de inicializar `path` otra vez usando el codigo arriba.

<h2>Crear marcas de tiempo para detecciones simuladas de movimientos</h2>

Los datos primarios que colecta ABISSMAL son marcas de tiempo que indican el momento en el tiempo cuando un sensor se activo y grabo movimiento. Muchas (pero no todas) de estas detecciones se pueden asignar a uno o mas animales que se movieron cerca de un sensor, por ejemplo, cuando un pajaro entra a un contenedor de nido a traves de una antena circular de RFID ("radio frequency identification") montado en la entrada del contenedor. En los siguientes trozos de codigo, vamos a generar datos simulados que representan datos de detecciones grabados por el sensor de RFID y tambien sensores de infrarrojo.

Digamos que estamos recolectando datos para 2 pajaros adultos a traves de sensores de ABISSMAL montados en un contenedor de nido. La antena de RFID esta montada en la entrada del contenedor, y un par de sensores de infrarrojo, el par "externo", esta montado en frente de la antena de RFID para capturar movimientos afuera de la entrada del contenedor. Un segundo par de sensores de infrarrojo, el par "interno", esta montado detras de la antena de RFID para capturar movimientos adentro de la entrada del contenedor. En este ejemplo simulado, el par externo de sensores infrarrojo va a activar cuando un pajaro entra al contenedor de nido, luego la antena de RFID, y luego el par interno de sensores infrarrojos. Cuando un pajaro sale del contenedor, el par interno de sensores infrarrojos deberia de activar primero, luego la antena de RFID, y luego el par externo de sensores infrarrojos.

TKTK continue translation

We will start by creating an object that holds simulated timestamps for the RFID antenna. This object will be called `rfid_ts`, and it will hold 4 timestamps in hours:minutes:seconds format. Each timestamp will be surrounded by double quotes to denote that we are using text or character string information here. 

These timestamps are combined inside of the same object by using the function `c()`. This function `c()` concatenates values separated by commas into a vector or list object. Above you created an object called `path` without using `c()`, but that object had a single value. Using `c()` allows you to combine multiple values into a vector that will have more than 1 element.
```{r}

# Create a vector of 4 RFID timestamps or 4 elements in HH:MM:SS format
rfid_ts <- c("10:00:00", "10:05:00", "11:00:00", "11:05:00")
  
```

You can check out the different properties of the `rfid_ts` object after it's made:
```{r}

rfid_ts # Run the object name to see its contents

is.vector(rfid_ts) # A binary value indicating that rfid_ts is a vector

class(rfid_ts) # A vector with data of type "character"

length(rfid_ts) # This vector has 4 elements

```

Let's continue by simulating 2 entrance and 2 exit movement events. We can choose timestamps for the infrared beam breakers that precede the RFID timestamps to simulate an entrance event, and infrared beam breaker timestamps that follow the RFID timestamps to simulate an exit event. We will offset detections from each sensor within the entrance and exit movement events by 1 second.
```{r}

# Simulate outer and inner beam breaker timestamps for an entrance, an exit, and then another entrance and exit
o_irbb_ts <- c("09:59:59", "10:05:01", "10:59:59", "11:05:01")
i_irbb_ts <- c("10:00:01", "10:04:59", "11:00:01", "11:04:59")

```

Birds will sometimes perch in the nest container entrance, and the sensors should pick up this behavior. You can add some perching events to the simulated dataset. Here we will simulate perching events for the RFID data only.
```{r}

rfid_ts <- c(rfid_ts, "08:00:00", "08:00:01", "08:00:02", "08:00:03", "11:30:00", "11:30:01", "11:30:02", "11:30:03", "11:30:04", "11:30:05")

```

In the code above, we modified the object `rfid_ts` by using the function `c()` to add 10 more timestamps to this vector, for a total of 14 elements. Check out the structure of the modified `rfid_ts` object using the function `glimpse()`:
```{r}

# See the object structure in a format that includes the data type ("chr"), the number of elements ([1:14]), and a preview of the first few elements
glimpse(rfid_ts)

```

Another important type of information to simulate is noise in the sensor detections. For instance, the RFID antenna can fail to detect an individual's passive integrated transponder (PIT) tag, and the infrared beam breakers can activate when birds leave nesting material hanging in the entrance of the container. In both of these cases, the infrared beam breakers should activate when the RFID antenna does not.
```{r}

# Simulate some RFID detection failures across both beam breaker pairs
# These RFID detection failures occurred over 2 additional simulated entrances and 2 additional simulated exits
o_irbb_ts <- c(o_irbb_ts, "06:05:05", "06:35:08", "07:15:40", "11:10:25")
i_irbb_ts <- c(i_irbb_ts, "06:05:04", "06:35:09", "07:15:41", "11:10:24")

glimpse(o_irbb_ts)
glimpse(i_irbb_ts)

# Simulate some stray detections for the outer beam breaker
o_irbb_ts <- c(o_irbb_ts, "09:45:01", "09:45:02", "09:45:03", "09:45:04", "09:45:05", "09:45:06", "09:45:07", "09:45:08", "09:45:09", "09:45:10", "09:45:11")

glimpse(o_irbb_ts)

```

You've created simulated, noisy datasets of detections of animal movements, but currently, these datasets are encoded across separate vector objects and do not have important metadata. For instance, some useful metadata includes information about the experimental replicate, the date, and for the RFID data, the unique alphanumeric code of each PIT tag detected by the RFID antenna. Some of this metadata is critically important for downstream analyses (e.g. the date and PIT tag codes).

Vectors are useful data structures in R, but one limitation of vectors is that you cannot combine different data types into the same object. You can try this out with the code below:
```{r}

# Create a vector with character, numeric, and binary data types
# Since you're not saving the output into an object, the result will print directly to the console
# You should see that all elements are forced to be character strings in double quotes
c("1", 1, TRUE, FALSE)

# Create a vector with numeric and binary data types
# You should see that all elements have been forced to numeric. The TRUE and FALSE binary values are converted to the underlying numeric values used to store binary information (TRUE is stored as 1, and FALSE is stored as 0)
c(1, 1, TRUE, FALSE)

```

When you try to combine character, numeric, and binary data types in the same vector, all vector elements will be forced to the character data type. A similar thing happens when you try to combine numeric and binary data types, but the vector is forced to numeric. Note that the binary values TRUE and FALSE are equivalent to the numeric values 1 and 0, respectively.

<h2>Create metadata vectors</h2>

For downstream analyses, you need to be able to create metadata of different data types, and then combine the primary data for each sensor (timestamps) with this metadata. Start by creating vectors of metadata for the RFID timestamps, including information about the experimental replicate, the date, and the PIT tag identity for each detection.

Make a vector of metadata about the experimental replicate. Here, instead of writing out the same information many times, you'll use the function `rep()` to repeat the information about the experimental replicate several times. In order to set the number of times that the experimental replicate information is repeated, you'll also use the function `length()` to get the length of the `rfid_ts` vector, and then feed that result to `rep()`. Creating a metadata vector that is the same length as the `rfid_ts` vector will be useful for combining these vectors into a single object later. 
```{r}

# The documentation tells us that rep() expects two arguments, `x` and `times`
?rep

# Create a vector with information about the experimental replicate
# The argument `x` contains the metadata information that will be repeated
# The argument `times` specifies the number of times that the information will be repeated
exp_rep <- rep(x = "Pair_01", times = length(rfid_ts))

glimpse(exp_rep)

# You can also run code without writing out the argument names, as long as the arguments are written in the same order that the function expects:
exp_rep <- rep("Pair_01", length(rfid_ts))

glimpse(exp_rep)

```

Using `times = length(rfid_ts)` is better practice than manually entering the length of `rfid_ts` (e.g. `times = 14`). When you manually enter a value for `times`, you're assuming that the `rfid_ts` has not changed within and between coding sessions, which can be a dangerous assumption. By using `times = length(rfid_ts)`, you are ensuring that the code above will create a vector of metadata the same length as `rfid_ts` regardless of whether you made additional modifications to `rfid_ts` above.

You can use a conditional statement to confirm that the vector of metadata is the same length as the vector of RFID timestamps. Conditional statements can be useful ways to check assumptions in your code, or to efficiently build new datasets and functions.
```{r}

# If this condition is met, then the result in the console should be "[1] TRUE"
length(rfid_ts) == length(exp_rep)

```

In the conditional statement above, you're using the symbols `==` to ask if the two vectors `rfid_ts` and `exp_rep` are the same length.

You can also use the symbols `!=` to ask if the two vectors are *not* the same length:
```{r}

# This statement should yield FALSE, because these vectors are the same length
length(rfid_ts) != length(exp_rep)

```

As an example, you can modify `rfid_ts` so that it is a different length than `exp_rep`. Below you'll see some different ways that you can filter out 4 elements of the vector `rfid_ts` so that this vector is 10 elements long.
```{r}

# You can use : to create a sequence of numbers from indices 5 to the length of rfid_ts
5:length(rfid_ts)

# You can also use the function `seq()` to create the same sequence of numeric indices
seq(from = 5, to = length(rfid_ts), by = 1)

# If you want to filter out non-consecutive elements, you can create a vector of indices with the function `c()`
c(1, 3, 5, 6, 8, 10, 11, 12, 13, 14)

# When you put any of the expressions above inside of square brackets after the object name you will pull out elements 5 to the length of rfid_ts and drop the first 4 elements
rfid_ts[5:length(rfid_ts)]

rfid_ts[seq(from = 5, to = length(rfid_ts), by = 1)]

rfid_ts[c(1, 3, 5, 6, 8, 10, 11, 12, 13, 14)]

# Finally, you can use any of the methods above to create a sequence of indices that you want to drop, and then use the `-` symbol inside of the square brackets to remove those indices. For instance:
rfid_ts[-c(1:4)] # the numbers must be wrapped in `c()` in order for this inverted filtering to work 

```

Next, you can check whether the modified version of `rfid_ts` is the same length as `exp_rep`:
```{r}

# This statement should yield TRUE, because these vectors are not the same length
length(rfid_ts[-c(1:4)]) != length(exp_rep)

```

<h2>Create data frames with primary data and metadata</h2>

For later analyses, it is really important to have the metadata accompany your primary data in the very same object. To combine the primary data and metadata for these simulated datasets, you can rely on an object called a "data frame". Data frames are similar to spreadsheets. They have 2 dimensions (rows and columns), and you can store multiple different types of data in the same data frame. You can also write out data frames to spreadsheets that exist as physical files on your computer.

When you try making a data frame with vectors of different length, you should get an error message stating that the two arguments provided have different numbers of rows:
```{r eval = FALSE}

sim_dats <- data.frame(exp_rep, rfid_ts[-c(1:4)])

# "Error in data.frame(exp_rep, rfid_ts[-c(1:4)]) : 
  # arguments imply differing number of rows: 14, 10"

```

Now use the full `exp_rep` and `rfid_ts` vectors to create the data frame, in which these vectors become columns of the data frame:
```{r}

sim_dats <- data.frame(exp_rep, rfid_ts)

glimpse(sim_dats)

```

When you check out the structure of the new data frame `sim_dats`, you can see that it has 14 rows and 2 columns. For each column (after the "$" symbol), you can see the column name (here `exp_rep` and `rfid_ts`), the type of data in each column (both columns are type "character"), and then a preview of the first values in each column.

You can change the column names by adding a new name and `=` before each vector. Here the vector `exp_rep` becomes the column `replicate` and the vector `rfid_ts` becomes the column `timestamps``
```{r}

sim_dats <- data.frame(replicate = exp_rep, timestamps = rfid_ts)

glimpse(sim_dats)

```

We can add additional metadata to this data frame now that it's been created. For instance, it would be useful to have information about the date when these timestamps were collected. You can start by adding a column for the year, which will be data type "dbl" (which stands for "double" or "numeric"):
```{r}

sim_dats <- cbind(sim_dats, rep(2023, length(rfid_ts)))

glimpse(sim_dats)  

```

The year column has a strange name when we add the new column using `cbind()`. The column name is the expression that you wrote inside of `cbind()`. You can use the function `names()` and square bracket indexing to change the strange column name to "year":
```{r}

# This function returns a vector of the column names of the data frame 
names(sim_dats)

# Use square bracket indexing and the function `ncol()` to find the last column name
ncol(sim_dats) # There are 3 columns in this data frame

# This expression gets you the name of the last column
names(sim_dats)[ncol(sim_dats)]

# Then you can overwrite the last column name with a new name
names(sim_dats)[ncol(sim_dats)] <- "year"

# Confirm that the name was changed correctly
names(sim_dats)
glimpse(sim_dats)

```

<h2>Create data frames using the tidyverse</h2>

In the chunk above, you used base R code to add a new column and to update the column name. It took more than a few lines of code to carry out these operations. You could reduce the amount of code needed for these steps by removing the lines included to check your work. But another way that you can reduce the amount of code you're writing for these operations is to use tidyverse notation and functions:
```{r}

# Make the data frame again with only 2 columns
sim_dats <- data.frame(replicate = exp_rep, timestamps = rfid_ts)

# Use the tidyverse to add the year as the 3rd column
sim_dats <- sim_dats %>% 
  dplyr::mutate(
    year = rep(2023, length(rfid_ts))
  )

glimpse(sim_dats)

```

You added a column for the year with the correct column name in fewer lines of code. In this tidyverse syntax, the symbol `%>%` represents a piping operation, in which you're using one object as the input for a subsequent operation. Here, you're using the object `sim_dats` at the input for the function `mutate()`, which you used to create the column `year`. 

The notation `dplyr::` before `mutate()` indicates that the function `mutate()` should be sourced from the package called `dplyr()`. Including the package name with 2 colons before a function name is important when there are multiple functions loaded in your global environment that have the same name. For instance, if you use other packages that also have functions called `mutate()`, and you don't specify which package you want to use, then you could end up with immediate errors (the code fails to run) or worse, you could run the wrong `mutate()` operation for a given analysis (which can lead to errors down the line that are harder to trace).

Piping operations can simplify the code that you write, so that you don't have to create as many intermediate objects. On the other hand, for the same reason it can take practice to troubleshoot piping operations, especially when they become very long. A useful way to check intermediate results within long piping operations is to include the function `glimpse()` between different piping steps:
```{r}

# Make the data frame again with only 2 columns
sim_dats <- data.frame(replicate = exp_rep, timestamps = rfid_ts)

# Use the tidyverse to add the year as the 3rd column
sim_dats %>%
  glimpse() %>% # See the structure of the first version sim_dats
  dplyr::mutate(
    # The nrow(.) expression means "get the number of rows for the current object", which in this case is sim_dats
    year = rep(2023, nrow(.))
  ) %>%
  glimpse() # See the structure of the latest sim_dats with the new column "year"

```

In the code above, you also learned a new way to repeat a value within a piping operation by using the notation '.' inside of a function, which means that you will perform the given operation (getting the number of rows) on the current object (`sim_dats`) piped into the expression (the `nrow()` function). 

You can use a similar rule of thumb to add two more columns for the month and day to the data frame:
```{r}
 
# Use the tidyverse to add the year as the 3rd column
sim_dats %>%
  glimpse() %>% # See the structure of the original sim_dats
  dplyr::mutate(
    year = 2023
  ) %>%
  glimpse() %>% # See the structure of the intermediate sim_dats with the new column year
  # Also add columns for the month and day
  dplyr::mutate(
    month = 08,
    day = 01
  ) %>% 
  glimpse() # See the structure of the final sim_dats with the additional new columns month and day

```

In the code above, you added 2 more numeric columns to the data frame and you did this without needing to use the `rep()` function. Since you piped an existing data frame into each `dplyr::mutate()` expression, the single value that you specified for each of the year, month, and day columns was automatically repeated to fill the total rows in the data frame. Specifying a single value for a new column can help reduce the amount of code that you write, but only when you truly want the same value repeated across all rows of a data frame.

You did not save these modifications that you made to `sim_dats` in an object, so the output was printed to the console only. In the next vignette, you will write out this data frame of simulated RFID and IRBB data to spreadsheets as physical files on your computer.