---
title: "Vignette 05: Process and Plot Data"
author: "Grace Smith-Vidaurre"
date: "2023-12-27"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

```

<h2>Vignette Overview and Learning Objectives</h2>

In this fifth vignette, you will begin using the simulated detections of animal movements in the ABISSMAL data processing and analysis workflow, including combining raw data across days and pre-processing the raw data. You will also make visualizations of the processed data. You will continue to use coding skills that you learned in the previous vignettes, and you will learn additional skills that include:

1. Accessing custom functions
2. Using custom functions
3. Making line graphs with ggplot

<h2>Load packages and your working directory path</h2>

```{r message = FALSE, warning = FALSE}

rm(list = ls()) # Clean global environment

library(tidyverse) # Load the set of tidyverse packages
library(data.table) # Load other packages that the ABISSMAL functions require

# Initialize an object with the path that is your working directory
path <- "/home/gsvidaurre/Desktop/ABISSMAL_vignettes"

```

<h2>Load ABISSMAL functions</h2>

The custom R functions available through the ABISSMAL GitHub repository are stored in physical files (extension .R) inside the local repository on your computer (which you should have downloaded in vignette 01). In order to start using the ABISSMAL functions, you need to load the physical .R files so that the functions are available in your global environment. In the code below, you will use the function `source()` to load 3 of the 5 main ABISSMAL functions, plus a script that holds a set of utility functions:
```{r}

# Load the function that combines raw data
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/combine_raw_data.R")

# Load the function that detects perching events in the raw data
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/detect_perching_events.R")

# Load the function that pre-processes raw data
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/preprocess_detections.R")

# Load a script with utility functions that each function above requires
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/utilities.R")

```

<h2>Access ABISSMAL function information</h2>

After running the lines of code above, you should see that a whole set of functions have been loaded into your global environment (check the `Environment` pane). Many of these functions start with `check_`, and those are utility functions. If you scroll down, you'll see that the three main functions above (`combine_raw_data`, `detect_perching_events`, `preprocess_detections`) are all loaded in your global environment. In the column to the right of the function names you can also see a preview of each function's arguments. 

To get more information about each of these three main functions, you can click the white square icon to the very right of each function in the `Environment` pane, or run the code `View(function_name)`. This will open the script for the given function a new tab in your Source pane. In each script for each function, you'll see lines of documentation starting with the symbols "`# @". You'll see the function name and description first, and then a description of each argument (parameter) for the function. If you keep scrolling down, you'll see a section with details about how the given function works, and the information that it returns. After the lines of documentation, you'll see the code that makes up the function itself. 

<h2>Combine raw data</h2>

Once you've loaded the ABISSMAL functions, you can start using the first function, `combine_raw_data()`, to combine data collected across days per sensor into a single spreadsheet per sensor. You'll start by combining raw data for the RFID sensor collected over different days into a single spreadsheet.

Here goes more information about the arguments you are supplying to the function below:

* `sensors` is a vector containing the labels of the sensors for which you want to combine raw data. Below you're specifying RFID as a single sensor

* `path` is your general working directory

* `data_dir` is the folder that holds data inside of your working directory

* `out_dir` is the folder where you want to save the combined raw data spreadsheet. The function will create this folder if it does not already exist

* `tz` is the timezone for converting timestamps to POSIXct format. The default is "America/New York", and you can check out the "Time zones" section in the documentation for DateTimeClasses in R for more information (`?DateTimeClasses`)

* `POSIXct_format` is a string containing the POSIX formatting information for how dates and timestamps should be combined into a single column. The default is to encode the year as a 4 digit number and the month and day as 2 digit numbers, separated by dashes. The date is followed by a space, then the 2-digit hour, minute, and decimal second (separated by colons)
```{r}

combine_raw_data(sensors = "RFID", path = path, data_dir = "Data", out_dir = "Data/raw_combined", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

You can check that `combine_raw_data()` wrote a single spreadsheet of raw RFID data to the new directory `raw_combined`:
```{r}

list.files(file.path(path, "Data/raw_combined"), pattern = ".csv$")

```

You can read this combined .csv file (called "combined_raw_data_RFID.csv") back into R to check out the structure of this spreadsheet:
```{r}

rfid_data <- read.csv(file.path(path, "Data/raw_combined", "combined_raw_data_RFID.csv"))

glimpse(rfid_data)

```

You should be able to see that there are some new columns created by the function, such as the column `data_type`. For this spreadsheet, the columns `sensor_id` and `data_type` contain the same information, but it's useful to have separate columns to keep track of the sensor id and type of sensor used to collect data when multiple sensors are used (e.g. two beam breaker pairs will each hace a unique identifier in `sensor_id`).

The function created a new timestamps column in POSIXct format for downsteam processing and analysis, but it kept the original timestamps column. The function also added columns to indicate the stage of data processing and the date that the raw data were combined. Finally, if you check the folders with the original raw RFID data, you'll see that the original spreadsheets per day were retained and not removed not overwritten.

You can also run `combine_raw_data()` with raw data from multiple sensors by supplying a vector with the sensor labels to the argument `sensors`:
```{r}

combine_raw_data(sensors = c("RFID", "IRBB"), path = path, data_dir = "Data", out_dir = "Data/raw_combined", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

The RFID data will be overwritten, and you should see an additional spreadsheet with the raw IRBB data:
```{r}

list.files(file.path(path, "Data/raw_combined"), pattern = ".csv$")

```

<h2>Detect perching events</h2>

You can use the combined raw data from different sensors in subsequent ABISSMAL functions to start making behavioral inferences from events in the movement detection datasets. For instance, you can detect perching events in the raw RFID data with the function `detect_perching_events()`. You can read more about each argument in the R script that contains this function.
```{r}

detect_perching_events(file_nm = "combined_raw_data_RFID.csv", threshold = 2, run_length = 2, sensor_id_col_nm = "sensor_id", timestamps_col_nm = "timestamp_ms", PIT_tag_col_nm = "PIT_tag", rfid_label = "RFID", general_metadata_cols = c("chamber_id", "sensor_id"), path = file.path(path, "Data"), data_dir = "raw_combined", out_dir = "processed", out_file_prefix = "perching_events", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

`detect_perching_events()` can only operate on a single file and sensor type at a time. The function will automatically create a new folder called "processed", and will save the .csv inside of that folder with perching events (if these were detected using the temporal threshold and run length above). 

When we simulated data in the third and fourth vignettes, you simulated perching events in the RFID dataset only. Did the code above recover those perching events?
```{r}

perching <- read.csv(file.path(path, "Data", "processed", "perching_events_RFID.csv"))

glimpse(perching)

```

`detect_perching_events()` identified 6 total perching events, which is the same number that you simulated in the previous vignette (two perching events per day). You can look at the values inside of data frame for more information about these perching events:
```{r}

# The timestamps when each perching event started
perching$perching_start

# The timestamps when each perching event ended
perching$perching_end

# The unique PIT tag identifier that tells you which individual was perched on the RFID antenna
perching$PIT_tag

```

You can also view the whole data frame in a separate pane:
```{r eval = FALSE}

View(perching)

```

The information above tells you that there were 2 perching events detected at 8:00 each day, and 2 perching events detected at 11:30 each day (as expected). The PIT tag for each individual was detected once each day, so there was 1 perching event performed by each individual on each day.

Detecting perching events is not a requirement in the ABISSMAL workflow, but it can be a useful step to obtain as much information from the raw data before some detections are dropped during pre-processing (see below).

<h2>Pre-process raw data</h2>

Once you've detected perching events in the raw data, you can move on to pre-processing the raw data itself with `preprocess_detections()`. The raw data can sometimes contain multiple detections separated by a short period of time (e.g. RFID detections when an individual is perching on the antenna), and these multiple detections can be a source of noise when you're trying to make behavioral inferences across data collected by multiple sensors. When the argument `mode` is set to "thin", `preprocess_detections()` removes detections separated by a very short period of time, and retains a reduced datasets of detections that still represents discrete movement events.

`preprocess_detections()` also operates on a single sensor type at a time:
```{r}

preprocess_detections(sensor = "RFID", timestamps_col_nm = "timestamp_ms", group_col_nm = "PIT_tag", mode = "thin", thin_threshold = 2, drop_tag = NULL, path = file.path(path, "Data"), data_dir = "raw_combined", out_dir = "processed", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

You should now see an additional .csv file called "pre_processed_data_RFID.csv" in the "processed" folder:
```{r}

list.files(file.path(path, "Data/processed"))

```

You can read this file into R to check out the data structure. You should see that there are fewer rows here compared to the spreadsheet of raw data:
```{r}

rfid_pp <- read.csv(file.path(path, "Data/processed/pre_processed_data_RFID.csv"))

glimpse(rfid_pp)

```

Next, you can pre-process the raw beam breaker data and check out the resulting .csv file:
```{r}

preprocess_detections(sensor = "IRBB", timestamps_col_nm = "timestamp_ms", group_col_nm = "sensor_id", mode = "thin", thin_threshold = 2, drop_tag = NULL, path = file.path(path, "Data"), data_dir = "raw_combined", out_dir = "processed", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

list.files(file.path(path, "Data/processed"))

irbb_pp <- read.csv(file.path(path, "Data/processed/pre_processed_data_IRBB.csv"))

glimpse(irbb_pp)

```

<h2>Plot RFID data in a barcode plot</h2>

Now that you've combined and processed the raw data per sensor, it's time to visualize these different datasets. Making visualizations as you write code is important for generating high-quality figures for publications and presentations, but also for checking your work.

In the code below, you'll learn how to use functions from the `ggplot2` package to make barcode style visualizations of the raw and pre-processed RFID detection datasets.

Start by reading in the raw and pre-processed RFID data, as well as the RFID perching events, and convert the timestamps to POSIX format.
```{r}

rfid_raw <- read.csv(file.path(path, "Data/raw_combined", "combined_raw_data_RFID.csv")) %>%
  # The timestamps must be converted to POSIX format every time that the data is read back into R for plotting
  dplyr::mutate(
    timestamp_ms = as.POSIXct(format(as.POSIXct(timestamp_ms, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6"))
  ) %>% 
  dplyr::arrange(-desc(timestamp_ms))

# You should see that the timestamp_ms is in "dttm" format, which means that the POSIX conversion was done correctly
glimpse(rfid_raw)

rfid_pp <- read.csv(file.path(path, "Data/processed/pre_processed_data_RFID.csv")) %>%
  # The timestamps must be converted to POSIX format every time that the data is read back into R for plotting
  dplyr::mutate(
    timestamp_ms = as.POSIXct(format(as.POSIXct(timestamp_ms, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6"))
  ) %>% 
  dplyr::arrange(-desc(timestamp_ms))

glimpse(rfid_pp)

rfid_perch <- read.csv(file.path(path, "Data/processed/perching_events_RFID.csv")) %>%
  # The start and end timestamps must be converted to POSIX format every time that the data is read back into R for plotting
  dplyr::mutate(
    perching_start = as.POSIXct(format(as.POSIXct(perching_start, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6")),
    perching_end = as.POSIXct(format(as.POSIXct(perching_end, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6"))
  ) %>% 
  dplyr::arrange(-desc(perching_start))

glimpse(rfid_perch)

```

Next, you can combine these datasets into a single data frame to facilitate combining them in the same plot. You will add a new column in order to identify the two different datasets.
```{r}

rfid_combined <- rfid_raw %>%
  dplyr::select(sensor_id, day, timestamp_ms) %>% 
  dplyr::mutate(
    dataset = "raw"
  ) %>% 
  bind_rows(
    rfid_pp %>%
      dplyr::select(sensor_id, day, timestamp_ms) %>% 
      dplyr::mutate(
        dataset = "pre-processed"
      ) 
  ) %>% 
  # Order the data frame by timestamps. The -desc() inside of arrange() means that the timestamps will be arranged in increasing order
  dplyr::arrange(-desc(timestamp_ms))

glimpse(rfid_combined)

```

You will start plotting the data using functions from `ggplot2`. This package is part of the `tidyverse`, but can also be installed and used separately. You can check out this [link](https://ggplot2.tidyverse.org/) for more resources to learn how to use ggplot to make different types of plots. These resources include sections of three different books with hands-on exercises at different levels, as well as an online course and a webinar.

Thde `ggplot2` package has a unique syntax for building plots, in which you start making a plot by calling the function `ggplot()`, and then add features by layering on other `ggplot2` functions with the `+` symbol.

If you call `ggplot()`, you'll see that the function immediately draws a blank plot in your Plots pane in RStudio.
```{r}

ggplot()

```

The plot will still remain blank even when you supply information about your data in order to set up plot aesthetics.
```{r}

ggplot(data = rfid_combined)

```

You need to add other aesthetics functions to this base layer of the plot in order to see your data. The functions that you use to layer aesthetics over the empty plot will depend on the type of plot that you want to make. For this example, you will make a barcode style plot, in which each timestamp is shown as a thin vertical line. Barcode plots can be useful visualizations when you're working with timestamps, since the most important information is contained in one dimension (time on the x-axis). If you were to summarize the number of timestamps recorded on each day, then you could create a line plot instead (see a plot in the next vignette).

You can layer the function `geom_segment()` over the base plot layer. `geom_segment()` allows you to add lines to a plot, and the lines can communicate information in one or two dimensions (width on the x-axis and height on the x-axis). In the plot you'll make below, you'll use `geom_segment()` to add lines to the plot that contain temporal information in one dimension (timestamps, information on the x-axis only).
```{r}

ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  )
  
```

In the code above, `geom_segment()` adds a vertical line segment to the plot for each detection in the full dataset. Using the argument `color` inside of `geom_segment()`, you told the function that these line segments should be colored by dataset by supplying the column name holding the dataset labels. The `color` argument must be inside of the `aes()` function in order for this color assignment by dataset to work correctly. 

The line segment colors are automatically assigned by ggplot using default colors, but you can change these colors using the function `scale_color_manual()`:
```{r}

ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen"))

```

The line segments are now colored with the new colors you specified. But in the legend, the pre-processed data shows up first. If you want to change the order of the dataset labels in the legend, and the colors assigned to them, you can modify the `dataset` column in the data frame that you used for plotting.

`ggplot` functions use a data type called `factors` for automated encoding of aesthetics, such as the color encoding you used above. Columns (or vectors) in factor format can look like `character` type columns, but R stores the unique values of each column as integers, and stores the unique character values in a property called "levels". You can change the order in which values of a column are plotted by changing the order of these levels of a factor column:
```{r}

# Change the column dataset to data type "factor"
# By specifying "raw" first in the argument levels, you are reordering the factor levels so that "raw" comes first 
rfid_combined <- rfid_combined %>% 
  dplyr::mutate(
    dataset = factor(dataset, levels = c("raw", "pre-processed"))
  )

# The dataset column is now type "fct", or "factor"
glimpse(rfid_combined)

# The levels of the factor column are ordered with "raw" first, rather than in alphabetical order
levels(rfid_combined$dataset)

```

After converting the `dataset` column to type "factor" amnd reordering the levels of the unique character values in this column, the unique values should appear in the correct order in the plot legend (not in alphabetical order). You should also see that the colors assigned to each dataset changed.
```{r}

ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen"))

```

In the plot that you just made, it's very difficult to discriminate between the lines for each dataset however. You can use the function `facet_wrap()` to split up the two datasets into different panels:
```{r}

ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen")) +
  
  # ~ is equivalent to "by", so that you're creating panels by the unique values in the column dataset
  facet_wrap(~ dataset, nrow = 2, strip.position = "left")

```

By faceting this plot, you split up the datasets into different panels but also lined them up on the x-axis so that it's easier to compare temporal patterns. 

From this point of view though, it's hard to see how the raw and pre-processed datasets differ. You can filter the data with `tidyverse` functions to plot the first 3 detections for each of the raw and pre-processed datasets:
```{r}

ggplot(data = rfid_combined %>%
         group_by(dataset) %>% 
         slice(1:2) %>% 
         ungroup()
       ) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen")) +
  
  # ~ is equivalent to "by", so that you're creating panels by the unique values in the column dataset
  facet_wrap(~ dataset, nrow = 2, strip.position = "left")

```

You should be able to see that the second timestamp in the raw data was dropped from the pre-processed dataset.

Next, you can add the perching data to the plot. This dataset doesn't have a single timestamps column, but rather has two columns that indicate the start and the end of each perching event, respectively. You can add this dataset to the full plot by using another function called `geom_segment()`. This function allows you to add lines to a plot, and the lines can communicate information in one or two dimensions (width on the x-axis and height on the x-axis). In the plot you'll make below, you'll use `geom_segment()` to add lines to the plot that contain temporal information in one dimension (on the x-axis only) about when perching events started and ended.
```{r}

ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen")) +
  
  # ~ is equivalent to "by", so that you're creating panels by the unique values in the column dataset
  facet_wrap(~ dataset, nrow = 2, strip.position = "left") +
  
  # Add the perching events to the plot
  geom_segment(
    data = rfid_perch,
    aes(x = perching_start, xend = perching_end, y = 1.2, yend = 1.5),
    color = "blue",
    linewidth = 0.3
  )

```

In the code above for `geom_segment()`, you specified that you wanted to add another dataset to the plot using the argument `data`. The arguments `x` and `y` determine where the beginning of each line segment is drawn on the x and y-axes, respectively. You also need to specify where you want the line segment to **end** on each axis. On the x-axis, by supplying the column name `perching_start` to the argument `x`, and `perching_end` to the argument `xend`, you're indicating that you want the line segment to start and end at the times when perching was inferred to start and end. On the y-axis, the numbers that you supplied to `y` and `yend` determined where the line segments for perching events were drawn, which was just above the other datasets. Note that the perching events were layered over each facet of the plot by default.

There are some additional changes you can make to this plot to make it more interpretable. You can change the position of the legend using the argument `legend.positon` inside of the general function `theme()`. Below you will also save the plot inside of an object, so that you don't have to write out all of the code over and over.
```{r}

gg <- ggplot(data = rfid_combined) +
  
  # Add a vertical line for each timestamp
  geom_segment(
    aes(x = timestamp_ms, y = 0, xend = timestamp_ms, yend = 1, color = dataset),
    linewidth = 0.3
  ) +
  
  scale_color_manual(values = c("orange", "darkgreen")) +
  
  # ~ is equivalent to "by", so that you're creating panels by the unique values in the column dataset
  facet_wrap(~ dataset, nrow = 2, strip.position = "left") +
  
  # Add the perching events to the plot
  geom_segment(
    data = rfid_perch,
    aes(x = perching_start, xend = perching_end, y = 1.2, yend = 1.5),
    color = "blue",
    linewidth = 0.3
  ) +
  
  theme(
    legend.position = "top"
  )

gg

```

You can make some minor adjustments to the plot to make it easier on the eyes, including changing the axis titles to be more informative, changing the background to be white, and removing the y-axis text and axis ticks(since the y-axis does not contain information for interpretation (the height of each segment does not reflect data that you want to interpret). You can also change the position of the legend so that it sits on top of the plot.
```{r}

gg <- gg +
  
  # Change the x and y axis labels
  xlab("Date and time") +
  
  # The y-axis does not contain information right now, so it can be blank
  ylab("") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top"
  )

gg

```

You can save plots that you make in R as physical image files. Below you'll use the function `ggsave()` to write the plot to an image file on your computer.
```{r}

gg

# Save the image file to your computer
ggsave(file.path(path, "raw_processed_perching.tiff"), width = 8, height = 6, units = "in", dpi = 300)

```

You can continue to modify minor aesthetics to this image file to create a high-quality figure for a publication. For instance, you can change the final size of the image file (`width`, `height`), as well as the resolution (`dpi`). You can also change the size of the text on each axis and the axis titles, or the legend position, as you play around with the final image size.