---
title: "Vignette 04: Save Data"
author: "Grace Smith-Vidaurre"
date: "2023-12-27"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

```

<h2>Vignette Overview and Learning Objectives</h2>

In this fourth vignette, you will write out spreadsheets of simulated detections of animal movements to your computer. You will continue to use coding skills that you learned in the previous vignettes, and you will learn additional skills that include:

1. Indexing and filtering data frames
2. Checking data frame structure with base R and the tidyverse
3. Saving objects from R as physical files on your computer 
4. Reading files from your computer into R
5. Writing and testing loops

<h2>Load packages and your working directory path</h2>

```{r message = FALSE, warning = FALSE}

rm(list = ls()) # Clean global environment

library(tidyverse) # Load the set of tidyverse packages

path <- "/home/gsvidaurre/Desktop/ABISSMAL_vignettes" # Initialize an object with the path that is your working directory

```

<h3>Create the simulated data</h3>

In the code below, you'll recreate the simulated RFID and beam breaker datasets that you learned to build in the previous vignette. Here the code from vignette 03 has been condensed into fewer chunks.
```{r}

# Create a vector of 4 RFID timestamps in HH:MM:SS format
rfid_ts <- c("10:00:00", "10:05:00", "11:00:00", "11:05:00")

# Add perching events to the RFID data
rfid_ts <- c(rfid_ts, "08:00:00", "08:00:01", "08:00:02", "08:00:03", "11:30:00", "11:30:01", "11:30:02", "11:30:03", "11:30:04", "11:30:05")

glimpse(rfid_ts)

```

```{r}

# For the beam breakers, simulate timestamps for an entrance, an exit, and then another entrance and exit
irbb_ts <- c("09:59:59", "10:05:01", "10:59:59", "11:05:01")

# Simulate some RFID detection failures for the beam breaker data
irbb_ts <- c(irbb_ts, "06:05:05", "06:35:08", "07:15:40", "11:10:25")

# Simulate some stray beam breaker detections
irbb_ts <- c(irbb_ts, "09:45:01", "09:45:02", "09:45:03", "09:45:04", "09:45:05", "09:45:06", "09:45:07", "09:45:08", "09:45:09", "09:45:10", "09:45:11")

glimpse(irbb_ts)

```

<h3>Simulate 2 days of RFID data collection</h3>

In the code below, you'll use combine the vector of RFID timestamps that you made above with metadata in a data frame. This metadata will include the year, month, and day, as well as a column with 2 unique PIT tag identifiers (1 for each simulated individual), and a column with the sensor type. You should recognize some of this code from the previous vignette:
```{r}

# Make a vector for the experimental replicate
exp_rep <- rep(x = "Nest_01", times = length(rfid_ts))

# Make the data frame with the experimental replicate metadata and the timestamps
sim_dats_rfid <- data.frame(chamber_id = exp_rep, timestamps = rfid_ts)

# Overwrite the data frame with the modified version that has columns for the year, month, and day
sim_dats_rfid <- sim_dats_rfid %>%
  dplyr::mutate(
    year = 2023
  ) %>%
  dplyr::mutate(
    month = 08,
    day = 01
  ) %>%
  # Allocate the first half of the detections to the first individual using nrow(.)/2, and the second half of the detctions to the second individual using nrow(.)/2 again. Both of the rep() expressions are combined into a single vector using the c() function
  dplyr::mutate(
    PIT_tag = c(rep("1357aabbcc", nrow(.)/2), rep("2468zzyyxx", nrow(.)/2))
  ) %>% 
  dplyr::mutate(
    sensor_id = "RFID"
  )

glimpse(sim_dats_rfid)

```

Next, you can use this data frame to simulate data collection over more than one day. To create more observations (rows) for two additional days, you can append rows from a modified copy of `sim_dats_rfid` to itself. 

In the code below, you're piping `sim_dats_rfid` into `bind_rows()`, which indicates that this is the original object to which you want to append new rows. Then the code inside of `bind_rows()` indicates the data frame (the new rows) that will be appended to `sim_dats_rfid`. In this case, the code inside of `bind_rows()` pipes `sim_dats_rfid` into `dplyr::mutate()`, which is a function that you're using to modify the `day` column to reflect a subsequent day of data collection.
```{r}

sim_dats_rfid <- sim_dats_rfid %>% 
  bind_rows(
    sim_dats_rfid %>% 
      dplyr::mutate(
        day = 02
      )
  )

glimpse(sim_dats_rfid) # Double the number of rows, looks good

```

<h3>Check data frame columns with base R</h3>

You checked that simulated data frame has data collected over two days looking at the structure of the new object. You can also check the unique values present in the column `day`. Below is a way of checking the unique values contained in a column of a data frame, using two different examples of base R notation for accessing columns:
```{r}

# The name of a data frame object, followed by a $ sign and then the name of a column allows you to pull out one column at a time from a data frame. A data frame column is a vector, so when you run this code you will see a vector of values print to the console
sim_dats_rfid$day

# You can also access a column in a data frame by using double square brackets after the name of the data frame, and placing the column name in quotes inside of the inner brackets
sim_dats_rfid[["day"]]

# You can use the function unique() to see the unique values of any vector, including a column of a data frame
unique(sim_dats_rfid$day) # Two days, looks good

unique(sim_dats_rfid[["day"]]) # Two days, looks good

```

<h3>Check data frame columns with the tidyverse</h3>

You can also check the unique values in a column using functions from the tidyverse. In the expression below, you're piping `sim_dats_rfid` into the function `pull()`, which lets you pull the column `day` out of the data frame as a vector, and then that vector is piped into the function `unique()` to check the unique values contained in the vector. The function `unique()` does not need an argument inside of the parentheses here because you're already piping the output that it needs directly into the function.
```{r}

# Two days, looks good
sim_dats_rfid %>% 
  pull(day) %>% 
  unique()

```

<h3>Simulate 2 days of beam breaker data collection</h3>

Next, repeat this process of creating a data frame with metadata for the infrared beam breaker dataset. Since the beam breakers do not collect unique individual identity information, you will add columns for the year, month, day, and sensor type. You will also simulate data collection for the beam breakers over the same two days as the RFID system.
```{r}

# Overwrite the vector exp_rep with a new vector the same length as irbb_ts
exp_rep <- rep(x = "Nest_01", times = length(irbb_ts))

sim_dats_irbb <- data.frame(chamber_id = exp_rep, timestamps = irbb_ts)

sim_dats_irbb <- sim_dats_irbb %>%
  dplyr::mutate(
    year = 2023,
    month = 08,
    day = 01,
    sensor_id = "Beam breakers"
  )

glimpse(sim_dats_irbb)

sim_dats_irbb <- sim_dats_irbb %>% 
  bind_rows(
    sim_dats_irbb %>% 
      dplyr::mutate(
        day = 02
      )
  )

glimpse(sim_dats_irbb) # Double the number of rows, looks good

# Two days, looks good
sim_dats_irbb %>% 
  pull(day) %>% 
  unique()

```

<h2>Save a data frame as a physical file</h2>

Data frames that you create and manipulate in R can be saved as physical files in your working directory. You have many different file type options for saving data frames, but I recommend using .csv format since this file type is compatible with R, Microsoft Word, and other software. You can use the function `write.csv()` to save data frames to .csv spreadsheets on your computer:
```{r eval = FALSE}

?write.csv

```

In order to write a physical file to your working directory, you need to tell R 1) where to save the file and 2) the name of the file that you want to create. You can pass both of these pieces of information to `write.csv()` by combining your working directory path and the file name through the function `file.path()`. For this example, you'll create a test file as you practice using `write.csv()`:
```{r}

# Create a custom file name by combining the path for your working directory with the file name that you want to write out
# The function file.path() will combine both pieces of information into a single file path
rfid_file <- file.path(path, "test_file.csv")
rfid_file

```

Next, you can pipe the data frame to `write.csv()` and specify additional information for creating the .csv file, such as whether you want to add a column of numeric row identifiers: 
```{r eval = FALSE}

sim_dats_rfid %>% 
  # Write out the data frame as a .csv spreadsheet. Do not include row names
  #The "." below means the function will use the object that is piped in, which here is the data frame sim_dats_rfid
  write.csv(x = ., file = rfid_file, row.names = FALSE)

```

As specified in the function documentation for `write.csv()`, the function will include column names in the resulting spreadsheet by default. The function will also not append new information to the .csv if it already exists. If you already created this file then it will be overwritten when you run the function again.

Next, you can check that `write.csv()` worked as expected by using `list.files()` to check the files in your working directory. In the second usage of `list.files()` below, you're using a new argument called `pattern` that makes it possible to customize a search. Using the argument `pattern` here is similar to searching for a specific word inside of a text document. Below, the dollar sign placed after ".csv" means that you're telling the function to search specifically for all files that *end* in the pattern ".csv".
```{r eval = FALSE}

# List all files in the given path
list.files(path)

# List only files that end in the pattern ".csv" in the given path
list.files(path, pattern = ".csv$")

```

<h3>Read in a spreadsheet</h3

You can read one these files back into R with the function `read.csv()`. In the code below, you're piping the output of `read.csv()` directly into `glimpse()` to check out the structure of the resulting data frame. The output of the code is printed to the console but is not saved inside of an object for later manipulation. Instead, you will be using a custom function from ABISSMAL that automatically reads in each spreadsheet.
```{r eval = FALSE}

read.csv(file.path(path, "test_file.csv")) %>% 
  glimpse()

```

Now that you've practiced using `write.csv()` and `read.csv()`, you can delete the temporary file that you created by feeding the `rfid_file` object to the function `file.remove()`.
```{r eval = FALSE}

rfid_file <- file.path(path, "test_file.csv")
rfid_file

file.remove(rfid_file)

```

<h2>Save simulated data for the ABISSMAL workflow</h2>

In the code below, you'll work through a pipeline to save the simulated data in the format and locations expected by the custom ABISSMAL functions. In order to use the ABISSMAL pipeline, the simulated raw data needs to be saved in a separate spreadsheet per sensor and day of data collection. These spreadsheets need to be saved inside of a folder per sensor as well. ABISSMAL will automatically save data in these ways when the tracking system is used to collect data from live animals.

<h3>Filter a data frame</h3>

In the code below, you'll practice how to use the function `dplyr::filter()` to filter rows of a data frame by day and then save a filtered data frame with `write.csv()`. To filter a data frame, you can use a conditional statement inside of the function `dplyr::filter()`:
```{r}

# Pipe the data frame into the filter() function
sim_dats_rfid %>% 
  # Filter the data frame by pulling out all rows in which the day column was equal to 1
  dplyr::filter(day == 1) %>%
  glimpse()

# Check that the filtering was done correctly by getting the unique values in the day column. Looks good
sim_dats_rfid %>% 
  dplyr::filter(day == 1) %>%
  pull(day) %>% 
  unique()

```

You can can similar results by inverting the conditional statement inside of `dplyr::filter()` to remove days that are not the first day of data collection.
```{r}

sim_dats_rfid %>% 
  # Filter the data frame by pulling out all rows in which the day column was not equal to 2
  dplyr::filter(day != 2) %>%
  glimpse()

# Check that the filtering was done correctly by getting the unique values in the day column. Looks good
sim_dats_rfid %>% 
  dplyr::filter(day != 2) %>%
  pull(day) %>% 
  unique()

```

Now that you've practiced filtering a data frame, you can combine this filtering step with writing out a .csv file that contains the filtered data frame with data collected on a single day.
```{r eval = FALSE}

# Make a new directory inside of your working directory for saving data
file.path(path, "Data") # Check the new path
dir.create(file.path(path, "Data")) # Create the new path

# Then make a new directory inside of the Data folder for the raw RFID data
file.path(path, "RFID") # Check the new path
dir.create(file.path(path, "Data", "RFID")) # Create the new path

# Make the new file name with your working directory path
# Make sure to specify that the spreadsheet will be saved inside of the new folder "RFID"
rfid_file <- file.path(path, "Data/RFID", "test.csv")
rfid_file

sim_dats_rfid %>% 
  dplyr::filter(day == 1) %>% 
  # Write out the data frame as a .csv spreadsheet. Do not include row names
  # Remember that the "." means the function will use the object that is piped in, which here is the data frame filtered by day 1
  write.csv(x = ., file = rfid_file, row.names = FALSE)

```

Check that the test file was created inside of the new RFID folder, and then delete it.
```{r eval = FALSE}

list.files(file.path(path, "Data/RFID"), pattern = ".csv$")

rfid_file <- file.path(path, "Data/RFID", "test.csv")
rfid_file

file.remove(rfid_file)

```

<h3>Practicing writing a loop</h3>

In order to write out a data frame for each day of data collection per sensor, you could repeat the code above 4 times (twice per sensor). But it's good to avoid repeating code over and over, since this can lead to messy scripts as well as a greater risk for errors in data processing and analysis. Whenever you need to run the same code many times, you can write a loop. Loops are an important coding skill and we'll work through building a loop step by step.

To start, you can practice building a loop with the function `lapply()`.
```{r}

?lapply

# Make a vector of the files to write out
files <- c(file.path(path, "Data/RFID", "test1.csv"), file.path(path, "Data/RFID", "test2.csv"))

files
length(files)

# The argument X specifies the number of times that the loop will run. In this case, X is a numeric vector from 1 to the length of the files vector and contains the numbers 1 and 2. Therefore the loop will run twice, and each value in X will be used in each loop iteration to write out one file at a time

# The argument FUN is a custom function, written using the function(x){} notation. All of the code inside of the curly brackets {} will be run for each iteration of the loop. The argument x inside function() is the iterating variable, or the variable that will take on a different value in each iteration of the loop

# In this loop, the iterating variable x will take on each value in the vector supplied to the X argument. This means that in the first loop iteration, x will take on the numeric value 1. In the second loop iteration, x will hold the numeric value 2. To test that this is true, you can run the loop below, which will print the value of x per iteration to the console
lapply(X = 1:length(files), FUN = function(x){
  
  x
  
})

```

The iterating variable of the loop, or `x` does not exist as an object outside of the function. So if you print `x` outside of the loop, no object will be found. If you created a object called `x` outside of the loop above, then you will see the contents of that object instead. This means that writing a loop and setting `x` as the iterating variable will not affect other lines of code outside of the loop.

The iterating variable of a function does not always need to be `x`, but can be another letter (`i`, `j`, `y`, `z`) or any combination of multiple letters and numbers, underscores, and periods (as long as the variable name starts with a letter).

A useful property of the iterating variable is that you can use it to index vectors, data frames, or other objects that you created outside of the loop. For instance, you can use `x` and square bracket indexing to print the name of each file that you want to save.
```{r}

lapply(X = 1:length(files), FUN = function(x){
  
  files[x]
  
})

```

You can also modify the code inside the custom function to save each of these files, by placing the expression `files[x]` inside of the `write.csv()` function and piping a data frame to `write.csv()`.
```{r eval = FALSE}

lapply(X = 1:length(files), FUN = function(x){
  
  sim_dats_rfid %>%
    write.csv(file = files[x], row.names = FALSE)
  
})

```

You should see 2 "NULL" outputs printed to the console if the loop runs correctly. When you check the contents of the nested RFID folder, you should see that both .csv files were written out:
```{r}

list.files(file.path(path, "Data/RFID"))

```

You just wrote out 2 spreadsheets using 1 loop, but you wrote out the same data frame to each spreadsheet. In order to write out a different data frame to each spreadsheet, you add the data frame filtering that you learned above to the loop. In the code below, you'll create another vector object called `days`, and then use the iterating variable to filter `sim_dats_rfid` by each day in `days`.
```{r eval = FALSE}

days <- c(1, 2)

lapply(X = 1:length(files), FUN = function(x){
  
  sim_dats_rfid %>%
    # Filter the data frame by one day at a time
    dplyr::filter(day == days[x]) %>% 
    write.csv(file = files[x], row.names = FALSE)
  
})

```

You can delete these files that you created for testing. In the code below, you're seeing another pattern searching example. The string that you passed to the argument `pattern` starts with the symbol `^`, which is a symbol that indicates you're searching for all files that *start* with the pattern "test". You're also telling the function to return the full location of each file along with the file name, so that `file.remove()` knows exactly where to look for each file.
```{r eval = FALSE}

rem_files <- list.files(file.path(path, "Data/RFID"), pattern = "^test", full.names = TRUE)
rem_files

file.remove(rem_files)

```

<h3>Use a loop to save RFID spreadsheets</h3>

Now you can put all of these pieces together and use the loop to write out a spreadsheet per day for the RFID sensor.
```{r eval = FALSE}

# Make a vector of the custom file names to write out
files <- c("RFID_simulated_Pair-01_2023_08_01.csv", "RFID_simulated_Pair-01_2023_08_02.csv")

# Add the file path for the correct directory
files <- file.path(path, "Data/RFID", files) 
files

# Make a vector of the days to write out (1 day per iteration of the loop)
days <- c(1, 2)
days

# You can drop the lapply() argument names, since you're supplying the arguments in the order that the function expects
invisible(lapply(1:length(files), function(x){
  
  sim_dats_rfid %>%
    # Filter the data frame by one day at a time
    dplyr::filter(day == days[x]) %>% 
    write.csv(file = files[x], row.names = FALSE)
  
}))

```

In the code above, you'll see one additional change that we made to the loop by wrapping it in the function `invisible()`. This function silences the output printed to the console that you saw several times above, in which the output of each `lapply()` iteration is enclosed in double square brackets, and then a single pair of square brackets. `lapply()` is a function that returns a list, and when the function is carried out correctly but there is not output to print, the function will return `NULL` values (indicating empty output). This is the expected behavior of `lapply()` for our purposes, because we used `lapply()` to write out physical files rather than to return output as an object in R. Since you can check that the function worked by running `list.files()`, using `invisible()` helps clean up the amount of text that you need to check in your console.

```{r eval = FALSE}

# Both new .csv files for each day of RFID data are present, looks good
list.files(file.path(path, "Data/RFID"))

```

You can remove these files.
```{r eval = FALSE}

files <- c("RFID_simulated_Pair-01_2023_08_01.csv", "RFID_simulated_Pair-01_2023_08_02.csv")

# Add the file path for the correct directory
files <- file.path(path, "Data/RFID", files) 
files

file.remove(files)

```

You can continue practicing loops by repeating this process for the infrared beam breaker dataset.

<h3>Write a nested loop to save spreadsheets</h3>

If you want to cut down on the amount of code that you're writing, you could write out both the RFID and beam breaker data in the same loop. First you'll need to create another directory for the beam breaker data:
```{r eval = FALSE}

dir.create(file.path(path, "Data", "IRBB"))

```

To carry out the file filtering and writing, you'll use a type of object called a `list`, and then use these lists inside of a nested loop:
```{r}

# Make a list of the custom file names to write out
files <- list(
  c("RFID_simulated_Pair-01_2023_08_01.csv", "RFID_simulated_Pair-01_2023_08_02.csv"),
  c("IRBB_simulated_Pair-01_2023_08_01.csv", "IRBB_simulated_Pair-01_2023_08_02.csv")
)

glimpse(files)

```

Lists are useful objects to use in R because they're very flexible. Unlike vectors, a single list can hold many different types of data. Unlike a data frame, the elements of a list do not need to be the same dimension. The elements of a list can also be different data structures or object types. For instance, lists can contain vectors, data frames, and other lists all inside of the same object. The list that you created above has 2 elements, and each element is a vector of 2 strings containing file names that you supplied using the function `c()`.

You can index lists in a manner that is similar to indexing vectors and data frames, but indexing lists can be done with single or double square brackets for different outcomes:
```{r}

# Using a single square bracket to filter a list returns the first list element in list format
files[1]
glimpse(files[1])

# Using double square brackets returns the first list element only, so it removes the list structure and shows the original data structure of that element (here a vector)
files[[1]]
glimpse(files[[1]])

```

Lists can also have named elements, which makes it possible to access elements by name.
```{r}

# Make a named list of the custom file names to write out
files <- list(
  `RFID` = c("RFID_simulated_Pair-01_2023_08_01.csv", "RFID_simulated_Pair-01_2023_08_02.csv"),
  `IRBB` = c("IRBB_simulated_Pair-01_2023_08_01.csv", "IRBB_simulated_Pair-01_2023_08_02.csv")
)

glimpse(files)

```

Access the list elements by name:
```{r}

# Using a single bracket returns the first element as a list, called "RFID"
files["RFID"]

# Using the dollar sign or double square brackets returns the first list element in its original format
files$RFID

files[["RFID"]]

```

Lists are useful data structures for nested loops. For instance, if you want to write out a spreadsheet per day per sensor type, you need a loop to 1) iterate over sensor types, and then a loop to 2) iterate over days per sensor. You can use lists to create nested data structures to supply to a nested loop to make sure that each layer of the loop runs in the way that you expect. For instance, the list called `files` reflects the nested loop that you need because files are listed first by sensor type (each element of the list) and then by date (each element of the vector inside each list element):
```{r}

# Make a vector of sensor labels
sensors <- c("RFID", "IRBB")

sensors

# Make a named list of the custom file names to write out
files <- list(
  `RFID` = c("RFID_simulated_Pair-01_2023_08_01.csv", "RFID_simulated_Pair-01_2023_08_02.csv"),
  `IRBB` = c("IRBB_simulated_Pair-01_2023_08_01.csv", "IRBB_simulated_Pair-01_2023_08_02.csv")
)

files

# Make a list of file paths per sensor that will be used inside of the loop
file_dirs <- list(
  `RFID` = file.path(path, "Data/RFID"),
  `IRBB` = file.path(path, "Data/IRBB") 
)

file_dirs

# Make a list of the days to write out for each sensor
# This could be a single vector rather than a list since you want to write out the same days per sensor, but a list is useful in case you wanted to change the days (and number of days) to write out per sensor
days <- list(
  `RFID` = c(1, 2),
  `IRBB` = c(1, 2)
)

days

# Finally, you need to make a list of the data frames that you want to write out. Since you already know how to filter data frames by day, that code can go inside of the loop, and you only need to specify a data frame per sensor type here
dats <- list(
  `RFID` = sim_dats_rfid,
  `IRBB` = sim_dats_irbb
)

glimpse(dats)

```

Once you've set up the data structures that you want to loop over, you can write out the nested loop. Since this is a complex loop structure, it can be helpful to test the loop with set values of each iterating variable (`x` and `y`). 

After writing out this loop, but before running it, you should test the code inside of each loop layer. To do this, you can set the values of each iterating variable outside of the loop and then run the code inside of each loop. This form of testing is equivalent to freezing the loop in time, so that you're seeing the output of the code for a single loop iteration (here, the first iteration for each loop layer).

To carry out this type of testing, you should run the code to freeze the iterating variables on the first iteration of each loop. Then, you should run the code inside of each loop, starting with the creation of `days_tmp`, then the indexing and filtering the data frame, and then the filtering of the file names. You should not run either of the lines with `lapply()` in them, since you don't want either loop to fully execute.

As you test the code inside of each loop, you should see that between the outer and inner loops, you're using the name of the sensor for the given outer loop iteration ("RFID" for the first iteration) to set up the days over which the inner loop will iterate (in `days_tmp`). Inside of the inner loop, you're filtering the list of data frames by sensor, and then by the days that you want per sensor. Finally, you're indexing the file name for the right sensor type and day with a combination of double and single bracket filtering on the list of file names:
```{r eval = FALSE}

# Testing
x <- 1
y <- 1

# Start by iterating over sensors
invisible(lapply(1:length(sensors), function(x){
  
  # Index the named list of days to get the right days per sensor
  # This indexing is important to set up the next loop correctly
  
  sensors[x] # This is a string with the sensor name
  
  # Place the string with the sensor name inside of double square brackets to extract the vector of days for that sensor 
  days_tmp <- days[[sensors[x]]]
  
  days_tmp
  
  # For each sensor, iterate over days
  lapply(1:length(days_tmp), function(y){
    
    # Get the data frame per sensor type using x inside of double square brackets to extract the given data frame from the list
    dats[[x]] %>%
      # Filter the data frame by one day at a time by using y to index the temporary vector of days (e.g. to extract a single element from that vector)
      dplyr::filter(day == days_tmp[y]) %>% 
      glimpse()
    
    # Use double bracket filtering to pull out the vector of file names for the given sensor from the overall list, then use y with single bracket filtering to pull a single file name from the resulting vector of names
    files[[x]]
    
    files[[x]][y]
    
    # You'll also combine the file name with the right path:
    file.path(file_dirs[[x]], files[[x]][y])
    
  })
  
}))

```

You should have a better sense now of how each loop operates over different data structures to carry out the operation that you want (writing out a single spreadsheet per sensor type and day). Next, you can modify the full loop to replace the lines written for testing with the final operations that you want to carry out: 
```{r eval = FALSE}

# Start by iterating over sensors
invisible(lapply(1:length(sensors), function(x){
  
  # Index the list of days to get the right days per sensor
  # This indexing is important to set up the next loop correctly
  days_tmp <- days[[sensors[x]]]
  
  # For each sensor, iterate over days
  lapply(1:length(days_tmp), function(y){
    
    # Get the data frame per sensor type using x and then filter by day using y
    dats[[x]] %>%
      # Filter the data frame by one day at a time
      dplyr::filter(day == days_tmp[y]) %>%
      # Use double and single bracket filtering to pull out the right file name
      write.csv(file = file.path(file_dirs[[x]], files[[x]][y]), row.names = FALSE)
    
  })
  
}))

```

This loop should have created 1 file per sensor and day in the correct directory per sensor. Check that these 4 files now exist inside each directory per sensor within your working directory.
```{r eval = FALSE}

list.files(file.path(path, "Data/RFID"))

list.files(file.path(path, "Data/IRBB"))

```

You learned more about filtering data frames, writing them out to spreadsheets, and writing loops in this vignette. In the next vignette, you'll use the spreadsheets of simulated RFID and IRBB data to start the ABISSMAL data processing and analysis workflow.