---
title: "Vignette 06: Finish Analysis Pipeline"
author: "Grace Smith-Vidaurre"
date: "2023-12-27"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

```

<h2>Vignette Overview and Learning Objectives</h2>

In this sixth vignette, you will finish using the simulated detections of animal movements in the ABISSMAL data processing and analysis workflow. You will detect clusters of detections that represent distinct movement events, and then score behavioral inferences about the movement events. You will also make plots to visualize these behavioral inferences. You will continue to use coding skills that you learned in the previous vignettes, and you will additionally learn how to build more complex visualizations with ggplot.

<h2>Load packages and your working directory path</h2>

```{r message = FALSE, warning = FALSE}

rm(list = ls()) # Clean global environment

library(tidyverse) # Load the set of tidyverse packages
library(data.table) # Load other packages that the ABISSMAL functions require

# Initialize an object with the path that is your working directory
path <- "/home/gsvidaurre/Desktop/ABISSMAL_vignettes"

```

<h2>Load ABISSMAL functions</h2>

```{r}

# Load the function that detects clusters in the pre-processed data
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/detect_clusters.R")

# Load the function that scores behavioral inferences about clusters
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/score_clusters.R")

# Load a script with utility functions that each function above requires
source("/home/gsvidaurre/Desktop/GitHub_repos/ABISSMAL/R/utilities.R")

```

<h2>Finish the ABISSMAL pipeline</h2>

Here you'll use the ABISSMAL function `detect_clusters()` to identify clusters of detections across sensor types (e.g. detections from different sensors that were recorded close together in time). 
```{r}

# The run length needs to be set to 1 in order to correctly detect detection clusters of length 2
detect_clusters(file_nms = c("pre_processed_data_RFID.csv", "pre_processed_data_IRBB.csv"), threshold = 2, run_length = 1, sensor_id_col_nm = "sensor_id", timestamps_col_nm = "timestamp_ms", PIT_tag_col_nm = "PIT_tag", rfid_label = "RFID", preproc_metadata_col_nms = c("thin_threshold_s", "data_stage", "date_pre_processed"), general_metadata_col_nms = c("chamber_id", "year", "month", "day"), path = path, data_dir = "Data/processed", out_dir = "Data/processed", out_file_nm = "detection_clusters.csv", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

Next, you'll score behavioral inferences about these detection clusters with the function `score_clusters()`. 
```{r}

score_clusters(file_nm = "detection_clusters.csv", rfid_label = "RFID", camera_label = NULL, outer_irbb_label = "Outer Beam Breakers", inner_irbb_label = "Inner Beam Breakers", video_metadata_col_nms = NULL, integrate_perching = TRUE, perching_dataset = "RFID", perching_prefix = "perching_events_", sensor_id_col_nm = "sensor_id", PIT_tag_col_nm = "PIT_tag", pixel_col_nm = NULL, video_width = NULL, video_height = NULL, integrate_preproc_video = FALSE, video_file_nm = NULL, timestamps_col_nm = NULL, path = path, data_dir = "Data/processed", out_dir = "Data/processed", out_file_nm = "scored_detectionClusters.csv", tz = "America/New York", POSIXct_format = "%Y-%m-%d %H:%M:%OS")

```

<h2>Check final results</h2>

Now that you finished running the ABISSMAL pipeline to detect and make inferences about movement events, you can check the final results.
```{r}

scored_clusters <- read.csv(file.path(path, "Data/processed", "scored_detectionClusters.csv")) %>% 
  # The timestamps must be converted to POSIX format every time that the data is read back into R for plotting
  dplyr::mutate(
    start = as.POSIXct(format(as.POSIXct(start, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6")),
    end = as.POSIXct(format(as.POSIXct(end, tz = "America/New York"), "%Y-%m-%d %H:%M:%OS6"))
  ) %>% 
  dplyr::arrange(-desc(start))

glimpse(scored_clusters)

```

How many entrances and exits were scored per day? In order to count the number of each of these events scored per day, you need to 1) make a new column with information about the day per timestamp, 2) drop rows with missing data for direction scored, 3) group the data frame by day and direction scored, and 4) count the number of rows per group.
```{r}

scored_clusters %>%
  # Extract the day from each timestamp and make a new column with this information
  dplyr::mutate(
    day = lubridate::day(start)
  ) %>% 
  # Drop rows with missing values for direction_scored
  # Here you're using the function is.na(), which will return TRUE when it finds an NA (missing) value in the given column. By placing the ! before this function, you're inverting the output, so that all TRUE values are converted to FALSE. As a result, dplyr::filter will drop all rows that return the value FALSE (e.g. all rows with missing values in the column direction_scored)
  dplyr::filter(!is.na(direction_scored)) %>%
  # Group the data frame by both columns for which you want to count rows
  group_by(day, direction_scored) %>% 
  # The number of rows here is the number of exits or entrances scored per day
  dplyr::summarise(
    n = n()
  )

```

Four entrance and exit events were scored per day. Does this line up with the anticipated number of exits and entrances per day? If you go back to the code where you created the simulated raw datasets in vignettes 3 and 4, you should see that you started by simulating 2 entrances and exits per day across the RFID and beam breaker datasets. Then, you added 2 more entrance and 2 more exit events per day when you simulated RFID detection failures (e.g. movements captured by the beam breakers only). `score_clusters()` detected the correct number of entrances and exits per day.

Next, check out the perching events. Start by filtering out all rows that were not scored as perching events (rows that have NA values in the column "perching_PIT_tag"). Then, select only the columns that have information that is useful to check: the PIT tag codes, as well as the start and end of each perching event.
```{r}

scored_clusters %>% 
  # Use a conditional statement with is.na() to retain only rows that have PIT tag codes associated with perching events
  dplyr::filter(!is.na(perching_PIT_tag)) %>%
  # Then select only the columns that you want to check visually
  dplyr::select(start, end, perching_PIT_tag)

```

As specified in vignettes 3 and 4, the first perching event each day was performed by the first individual (PIT tag 1357aabbcc), and the second was attributed to the second individual (PIT tag 2468zzyyxx).

How many movement events that were not perching events were attributed to each individual?
```{r}

scored_clusters %>%
  # Extract the day from each timestamp and make a new column with this information
  dplyr::mutate(
    day = lubridate::day(start)
  ) %>%
  # Use a conditional statement with is.na() to retain only rows that have PIT tag codes not associated with perching events
  # Here you're combining two conditional statements, which allows you to search for rows in the column individual_initiated that have PIT tag codes, but that also do not have a sensor label in the column perching_sensor (e.g. movement events that were not perching events)
  dplyr::filter(!is.na(individual_initiated) & is.na(perching_sensor)) %>%
  group_by(day, individual_initiated) %>% 
  dplyr::summarise(
    n = n()
  )

```

As you can see, 4 movement events that were not perching events were attributed to the first individual (PIT tag code 1357aabbcc) on each day. This is exactly what we expected when creating the simulated dataset (see vignettes 3 and 4). More non-perching movement events were detected across these days, but those events were not always captured by the RFID antenna (e.g. the simulated RFID detection failures that were captured by the beam breakers only). The beam breakers captured those movement events but cannot record individual identity.

<h2>Build complex barcode plot</h2>

Now that you checked out the final results, you can visualize them. In the code below, you'll learn how to build a barcode plot that is more complex than the plot you made in vignette 5. For this plot, it would be helpful to visualize 3 types of behavioral inferences or types of information over time: the direction of movement (when available), the identity of the individual (when available), and perching events.

You can start building the plot by adding vertical line segments for non-perching movement events. The color of each line will indicate the individual identity, or whether individual identity could not be assigned. The line type of each line will indicate the direction of movement, or if direction could not be scored.

In order to encode colors and line types correctly in the plot, you need to modify the data frame of final results to convert NAs in the two associated columns to more meaningful information. For instance, when individual identity is missing, it would be useful to convert the associated NA value to "unassigned". In the code below, you will use the function `is.na()` inside of `ifelse()` conditional statements to change the NA values inside the columns individual_initiated and direction_scored to more useful information for plotting purposes.

First you can practice using `is.na()` inside of `ifelse()` to create a new vector. In the code below, you're providing the conditional statement that you want to test (here you're testing whether the day is the first day of the month), the value you want to add to the vector if the condition is true ("Day 1" of data collection), and then the value that you want to add if the condition is false ("Day 2" of data collection, since there are only 2 dates in each dataset). 
```{r}

# is.na() returns TRUE when it encounters NAs inside of a vector (or column), and FALSE when a value inside of the vector is not NA
is.na(scored_clusters$individual_initiated)

# All NAs in this column are converted to "unassigned", but all other values were not changed
ifelse(test = is.na(scored_clusters$individual_initiated), yes = "unassigned", no = scored_clusters$individual_initiated)

```

Now you can use `is.na()` inside of `ifelse` statements to modify columns in the data frame.
```{r}

scored_clusters_gg <- scored_clusters %>% 
  dplyr::mutate(
    # If this column has an NA value, then convert it to "unassigned"
    # Else, do not change the given value
    individual_initiated = ifelse(is.na(individual_initiated), "unassigned", individual_initiated),
    # Repeat this process for the direction_scored column but with a different value
    # Also, in the conditional statement below, you added is.na(perching_sensor) to only convert values of direction_scored when they were not labeled as perching events
    direction_scored = ifelse(is.na(direction_scored) & is.na(perching_sensor), "not scored", direction_scored)
  ) %>% 
  # Next, you'll convert each of these columns to type factor and arrange the levels in order for plotting
  dplyr::mutate(
    individual_initiated = factor(individual_initiated, levels = c("1357aabbcc", "2468zzyyxx", "unassigned")),
    direction_scored = factor(direction_scored, levels = c("entrance", "exit", "not scored"))
  )

# Check the resulting changes using the function distinct() to see all of the unique values for both columns that were modified
# The NA values in the direction_scored column are expected since these refer to perching events
scored_clusters_gg %>% 
  distinct(individual_initiated, direction_scored)

```

You can use this modified data frame to build the plot. In the code below, you'll add lines colored by individual identity and with line types encoding the direction of movement. First you'll specify the plot aesthetics:
```{r}

# Colors will be encoded in the same order as the levels of the column individual_initiated, so orange will encode "1357aabbcc"
levels(scored_clusters_gg$individual_initiated)
cols <- c("orange", "darkgreen", "black")

# Line types will be encoded in the same order as the levels of the column direction_scored, so dotted will encode "not scored"
levels(scored_clusters_gg$direction_scored)
ltys <- c("solid", "longdash", "dotted")
  
```

Then you can add lines to the plot by individual identity. Here, you're splitting up the `geom_segment()` calls by unique values in the individual_initiated column. You only added segments for the first individual and all non-perching events that were not assigned to an individual, because from checking the results above, you know that no non-perching events were assigned to the second individual.
```{r}

ggplot() +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg %>% 
      dplyr::filter(individual_initiated == "1357aabbcc"),
    aes(x = start, y = 0, xend = end, yend = 0.5, linetype = direction_scored),
    color = "orange",
    linewidth = 0.3
  ) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg %>% 
      dplyr::filter(individual_initiated == "unassigned"),
    aes(x = start, y = 0.6, xend = end, yend = 1.1, linetype = direction_scored),
    color = "black",
    linewidth = 0.3
  ) +
  
  # Add the custom linetype values to this plot
  scale_linetype_manual(values = ltys) +
  
  # Remove the y-axis label
  ylab("") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  # Add an argument to change where the legend is located in the plot
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top"
  )
  
```

To make this plot, you vertically offset the line segments for the first individual and those that were unassigned. This vertical offset makes it easier to visually compare patterns over time. You created this vertical offset by changing the values used for `y`
and `yend` in the second `geom_segment()` layer, so that those line segments would start higher than the end of the first set of segments. 

You can make some additional modifications that would help make this plot easier to interpret. First, the panel or facet labels on the left could be changed from the date to the day of data collection (e.g. "Day 1"). The x-axis labels could also be changed to show only the time of day (remove the month and day), and there could also be more labels available (e.g. a label each half hour).

You can start by modifying the data frame to add information about the day of data collection. You'll create this new column by relying on conditional statements with the `ifelse()` function, since there are only 3 days of data collection that need recoding.
```{r}

# Create a new column in the raw data for the date of data collection
scored_clusters_gg2 <- scored_clusters_gg %>% 
  # First you need to create a column with information about the day
  dplyr::mutate(
    day = lubridate::day(start)
  ) %>% 
  dplyr::mutate(
    # Then recode the label for each day and save this in a new column
    day_label = ifelse(day == 1, "Day 1", day), # Here the last argument is day because the column day_label does not exist yet
    day_label = ifelse(day == 2, "Day 2", day_label),
    day_label = ifelse(day == 3, "Day 3", day_label)
  )

# Looks good  
glimpse(scored_clusters_gg2)

scored_clusters_gg2 %>% 
  distinct(day_label)

```

Now you can update the code to include the new date labels:
```{r}

# Add the data frame as the default dataset for the base layer plot, so that the facet_wrap() layer below has data to plot
ggplot(data = scored_clusters_gg) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg2 %>% 
      dplyr::filter(individual_initiated == "1357aabbcc"),
    aes(x = start, y = 0, xend = end, yend = 0.5, linetype = direction_scored),
    color = "orange",
    linewidth = 0.3
  ) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg2 %>% 
      dplyr::filter(individual_initiated == "unassigned"),
    aes(x = start, y = 0.6, xend = end, yend = 1.1, linetype = direction_scored),
    color = "black",
    linewidth = 0.3
  ) +
  
  # Add the custom linetype values to this plot
  scale_linetype_manual(values = ltys) +
  
  # Remove the y-axis label
  ylab("") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  # Add an argument to change where the legend is located in the plot
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top"
  ) +
  
  # Facet the plot by day (e.g. create a panel per day)
  # Use the new day labels here
  facet_wrap(~ day_label, nrow = 3, strip.position = "left")

```

Now that information about the day of data collection has been moved to the facet labels, you need to fix the x-axis labels. The plot will be more interpretable if you can line up the timestamps by hour and minute for a direct comparison across days.  

To line up the timestamps across days, you will need to update the format of the columns that hold timestamps. The code to convert the timestamps to a different format is nested and repetitive, but the timestamp conversion will be performed correctly. When you prompt the R to convert the timestamps to hours, minutes, and seconds only, R adds a default year, month, and day beforehand (likely the current date that you rant the code). This is expected, and it is not an error, but rather makes it possible for the timestamps to align correctly over days in the plot (since R sees all timestamps occurring on a single day).
```{r}

scored_clusters_gg3 <- scored_clusters_gg2 %>% 
  dplyr::mutate(
    start_gg = as.POSIXct(strptime(format(as.POSIXct(start), "%H:%M:%S"), format = "%H:%M:%S")),
    end_gg = as.POSIXct(strptime(format(as.POSIXct(end), "%H:%M:%S"), format = "%H:%M:%S"))
  )
  
# You'll see that a new year, month, and day were appended to the updated timestamps, but this is expected (see above)
glimpse(scored_clusters_gg3)

```

Now you can update the plotting code to change the aesthetics of the x-axis using the function `scale_x_datetime()` to specify that you want x-axis labels every half hour. Also add an x-axis title, make the vertical line segments thicker, and remove the y-axis grid lines inside of each panel:
```{r}

# Update the dataset in the base layer of the plot, which will be used by all layers that do not have a specific data argument themselves
ggplot(data = scored_clusters_gg3) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "1357aabbcc"),
    aes(x = start_gg, y = 0, xend = end_gg, yend = 0.5, linetype = direction_scored),
    color = "orange",
    linewidth = 0.5
  ) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "unassigned"),
    aes(x = start_gg, y = 0.6, xend = end_gg, yend = 1.1, linetype = direction_scored),
    color = "black",
    linewidth = 0.5
  ) +
  
  # Add the custom linetype values to this plot
  scale_linetype_manual(values = ltys) +
  
  # Remove the y-axis label
  ylab("") +
  
  # Add a y-axis label
  xlab("Time") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  # Add an argument to change where the legend is located in the plot
  # Then remove the y-axis grid lines (major and minor) inside of each panel
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  
  # Facet the plot by day (e.g. create a panel per day)
  # Use the new day labels here
  facet_wrap(~ day_label, nrow = 3, strip.position = "left") +

  # Change the aesthetics of the x-axis labels
  scale_x_datetime(
    date_breaks = "30 mins",
    date_labels = "%H:%M"
  )

```

There's still one more important piece of information to add to the plot: the perching events. You'll add these using `geom_segment()`, but you'll make these line segments short and thick, so that they appear more like dots on the plot:
```{r}

ggplot(data = scored_clusters_gg3) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "1357aabbcc"),
    aes(x = start_gg, y = 0, xend = end_gg, yend = 0.5, linetype = direction_scored),
    color = "orange",
    linewidth = 0.5
  ) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "unassigned"),
    aes(x = start_gg, y = 0.6, xend = end_gg, yend = 1.1, linetype = direction_scored),
    color = "black",
    linewidth = 0.5
  ) +
  
  # Add the custom linetype values to this plot
  scale_linetype_manual(values = ltys) +
  
  # Remove the y-axis label
  ylab("") +
  
  # Add a y-axis label
  xlab("Time") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  # Add an argument to change where the legend is located in the plot
  # Remove the y-axis grid lines (major and minor) from each panel
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  
  # Facet the plot by day (e.g. create a panel per day)
  # Use the new day labels here
  facet_wrap(~ day_label, nrow = 3, strip.position = "left") +

  # Change the aesthetics of the x-axis labels
  scale_x_datetime(
    date_breaks = "30 mins",
    date_labels = "%H:%M"
  ) +
  
  # Add the perching events as rounded segments, and now encode color through the column individual_initiated
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(!is.na(perching_sensor)),
    aes(x = start_gg, y = 1.2, xend = end_gg, yend = 1.2, color = individual_initiated),
    linewidth = 2, lineend = "round"
  ) +
  
  # Add the custom colors
  scale_color_manual(values = cols)

```

You should see that a color legend appears at the top of the plot with the addition of the new `geom_segment()` layer. Both legends are a little messy. You can improve the plot by updating the title of each legend, increasing the size of the legend text, and reducing the white space between the plot and the legend. To modify the legend titles, you'll rely on the functions `guides()` and `guide_legend()`. To increase the legend text size, you'll use the argument `legend.text` inside of the function `theme()`. To reduce white space between the plot and legend, you'll use the argument `legend.margin` inside of the function `theme()`.
```{r}

gg <- ggplot(data = scored_clusters_gg3) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "1357aabbcc"),
    aes(x = start_gg, y = 0, xend = end_gg, yend = 0.5, linetype = direction_scored),
    color = "orange",
    linewidth = 0.5
  ) +
  
  # Add a vertical line for each non-perching event assigned to the first individual
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(individual_initiated == "unassigned"),
    aes(x = start_gg, y = 0.6, xend = end_gg, yend = 1.1, linetype = direction_scored),
    color = "black",
    linewidth = 0.5
  ) +
  
  # Add the custom linetype values to this plot
  scale_linetype_manual(values = ltys) +
  
  # Remove the y-axis label
  ylab("") +
  
  # Add a y-axis label
  xlab("Time") +
  
  # Use this function to convert the plot background to black and white
  theme_bw() +
  
  # Use aesthetics functions to remove the y-axis labels and ticks
  # Add an argument to change where the legend is located in the plot
  # Remove the y-axis grid lines (major and minor) from each panel
  # Increase the legend text size
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.text = element_text(size = 10),
    legend.margin = margin(-1, -1, -1, -1, unit = "pt")
  ) +
  
  # Facet the plot by day (e.g. create a panel per day)
  # Use the new day labels here
  facet_wrap(~ day_label, nrow = 3, strip.position = "left") +

  # Change the aesthetics of the x-axis labels
  scale_x_datetime(
    date_breaks = "30 mins",
    date_labels = "%H:%M"
  ) +
  
  # Add the perching events as rounded segments, and now encode color through the column individual_initiated
  geom_segment(
    data = scored_clusters_gg3 %>% 
      dplyr::filter(!is.na(perching_sensor)),
    aes(x = start_gg, y = 1.2, xend = end_gg, yend = 1.2, color = individual_initiated),
    linewidth = 2, lineend = "round"
  ) +
  
  # Add the custom colors
  scale_color_manual(values = cols) +
  
  guides(
    linetype = guide_legend(title = "Direction"),
    color = guide_legend(title = "Individual")
  )

gg

```

Finally, you can save this plot as an image file:
```{r}

gg

# Save the image file to your computer
ggsave(file.path(path, "behavioral_inferences.tiff"), width = 8, height = 6, units = "in", dpi = 300)

```

You can continue to modify minor aesthetics to this image file to create a high-quality figure for a publication. For instance, you can change the final size of the image file (`width`, `height`), as well as the resolution (`dpi`). You can also change the size of the text on each axis and the axis titles, or the legend position, as you play around with the final image size.